Fisher’s Method

### 2018-07-20 Fishers_method/Fishers_method_short.Rmd
minor edits

We believe that Fisher's test is not well suited to assess replication. In the event that you do reject $H_0$, this result only tells you that at least one study was a false negative, but it does not till you *which* study did in fact have a true effect. Even still, it provides no information about the size or direction of that true non-zero effect and whether or not it replicates the original finding. Presumably, a finding of "no false negatives" would be most informative in assessing replication in this scenario, but this can never be validly concluded from Fisher's method since that would require concluding the null hypothesis. While it is never advised to conduct a test in order to conclude the null hypothesis, this switched framework is especially problematic when the test is underpowered to reject $H_0$ because the Type II error rate will be large.

Even though the OSC was able to reject their null hypothesis ($X^2=155.83, p=0.048$), we think that in general Fisher's method is underpowered to answer the question at hand. We hypothesize that it requires many and possibly large non-null effects in order to skew the distribution of p-values enough to reject Fisher's null hypothesis. In the scenario where a researcher is combining the results of $k$ studies of the same treatment to test if an overall treatment effect $\Theta=0$, this type of conservative test may be appropriate. In the OSC scenario, however, since the 64 studies are not testing the same treatment effect, and one $\theta_j$ has no bearing on the 63 other $\theta_j's$, the presumed goal would be to detect if there are *any* false negatives among the replicate studies.

@@

For simplicity of interpretation but without loss of generality, we will work with the standardized scale of Cohen's d, defined as $\delta_j=\frac{\theta_j}{\sigma_j}$, where $\theta_j$ is the mean difference between the treatment and control groups in study $j$, and $\sigma_j$ is the known and equal variance among the treatment and control populations. Assuming equal sample sizes in the treatment and control groups within study $j$ (that is, let $n_j^t=n_j^c=n_j$), $\delta_j$ is estimated by $d_j\sim N(\delta_j,\frac{2}{n_j})$.[^2] Under this framework, Fisher's method can be represented in Cohen's d as testing the hypotheses

@@

We first consider the power of Fisher's method when $n_j=76/2=38$ for all $j$, because 76 is the median sample size in the OSC dataset, and we are assuming equal sample sizes in the treatment and control groups. The results of the power simulations for median sample size are given in Table 1.[^3] Note that 0.2, 0.5, and 0.8 correspond to small, medium, and large effects sizes on the scale of Cohen's d. As shown in the last column of Table 1, we find that there needs to be 10 large effects in order to achieve close to the standard 80% power (power=0.7743). If there is only one large effect, Fisher's test only has approximately 8% power to reject. Even when all 64 studies have small effects, Fisher's method only has 70% power to reject.

@@

In order to consider the power of Fisher's test under a "best-case scenario" in this dataset, we sort the 64 sample sizes and let the non-null effects be from the studies with the largest sample sizes first. That is, if there is just one non-null effect we let it be from the largest study; if there are two non-null effects we let them be from the two largest studies, etc.[^4] As shown in the first row of Table 2, even when a study with a very large sample size has a large effect, Fisher's method has less than 10% power to detect it. When the non-null effects come from the studies with the largest sample sizes, about half of the studies need to have $\delta=0.2$ in order to achieve approximately 80% power (power=0.8340).

Table: Power of Fisher's method given large sample sizes for varying $\delta$ and true # of non-null effects (i.e. "Best-case scenario")
```{r, echo=FALSE}
table <- readRDS("./power_large_n.RDS")
kable(table,format="markdown",digits=4,align=c('l','r','r','r'),col.names=c("# of non-null effects","$\\delta=0.2$","$\\delta=0.5$","$\\delta=0.8$"),caption="Power to reject H0 with large n")
```
[^4]:Note that because we are working with a set of replicate studies which found a p-value greater than 0.05, pairing large effects with very large sample sizes is not realistic, and therefore we begin the simulations with the largest $n_j$ among the studies for which the power is at most 99.99% to detect the given $\delta_j$. The largest sample sizes used for $\delta_j=0.2; 0.5$; and $0.8$ were $n_j=745; 159;$ and $100$ respectively. See Appendix A.3 for further discussion.

>>>>

We believe that Fisher's test is not well suited to assess replication. In the event that you do reject $H_0$, this result only tells you that at least one study was a false negative, but it does not tell you *which* study did in fact have a true non-zero effect. Even still, it provides no information about the size or direction of that non-zero effect and whether or not it replicates the original finding. Presumably, a finding of "no false negatives" would be most informative in assessing replication in this scenario, but this can never be validly concluded from Fisher's method since that would require concluding the null hypothesis. While it is never advised to conduct a test in order to conclude the null hypothesis, this switched framework is especially problematic when the test is underpowered to reject $H_0$ because the Type II error rate will be large.

Even though the OSC was able to reject their null hypothesis ($X^2=155.83, p=0.048$), we think that in general Fisher's method is underpowered to answer the question at hand. We hypothesize that it requires many and possibly large non-null effects in order to skew the distribution of p-values enough to reject Fisher's null hypothesis. In the scenario where a researcher is combining the results of $k$ studies of the same treatment to test if an overall treatment effect $\Theta=0$, this type of conservative test may be appropriate. In the OSC scenario, however, since the 64 studies are not testing the same treatment effect, and one $\theta_j$ has no bearing on the 63 other $\theta_j$'s, the presumed goal would be to detect if there are *any* false negatives among the replicate studies.

@@

For simplicity of interpretation but without loss of generality, we will work with treatment effects on the standardized scale of Cohen's d, defined as $\delta_j=\frac{\theta_j}{\sigma_j}$, where $\theta_j$ is the mean difference between the treatment and control groups in study $j$, and $\sigma_j$ is the known and equal variance among the treatment and control populations. Assuming equal sample sizes in the treatment and control groups within study $j$ (that is, let $n_j^t=n_j^c=n_j$), $\delta_j$ is estimated by $d_j\sim N(\delta_j,\frac{2}{n_j})$.[^2] Under this framework, Fisher's method can be represented in Cohen's d as testing the hypotheses

@@

We first consider the power of Fisher's method when $n_j=76/2=38$ for all $j$, because 76 is the median sample size in the OSC dataset, and we are assuming equal sample sizes in the treatment and control groups. The results of the power simulations for median sample size are given in Table 1.[^3] Note that 0.2, 0.5, and 0.8 correspond to small, medium, and large effects sizes on the scale of Cohen's d. As shown in the last column of Table 1, we find that there need to be 10 large effects in order to achieve close to the standard 80% power (power=0.7743). If there is only one large effect, Fisher's test only has approximately 8% power to reject. Even when all 64 studies have small effects, Fisher's method has just 70% power to reject.

@@

In order to consider the power of Fisher's test under a "best-case scenario" in this dataset, we sort the 64 sample sizes and let the non-null effects be from the studies with the largest sample sizes first. That is, if there is just one non-null effect we let it be from the largest study; if there are two non-null effects we let them be from the two largest studies, etc.[^4][^5] As shown in the first row of Table 2, even when a study with a very large sample size has a large effect, Fisher's method has less than 10% power to detect it. When the non-null effects come from the studies with the largest sample sizes, about half of the studies need to have $\delta=0.2$ in order to achieve approximately 80% power (power=0.8340).

Table: Power of Fisher's method given large sample sizes for varying $\delta$ and true # of non-null effects (i.e. "Best-case scenario")
```{r, echo=FALSE}
table <- readRDS("./power_large_n.RDS")
kable(table,format="markdown",digits=4,align=c('l','r','r','r'),col.names=c("# of non-null effects","$\\delta=0.2$","$\\delta=0.5$","$\\delta=0.8$"),caption="Power to reject H0 with large n")
```
[^4]:This provides a "best-case scenario" because large $n_j$'s lead to smaller $p_j$'s, which in turn result in a larger test statistic $X^2$ and greater likelihood of rejecting $H_0$ (i.e. higher power).

[^5]:Note that because we are working with a set of replicate studies which found a p-value greater than 0.05, pairing large effects with very large sample sizes is not realistic, and therefore we begin the simulations with the largest $n_j$ among the studies for which the power is at most 99.99% to detect the given $\delta_j$. The largest sample sizes used for $\delta_j=0.2; 0.5$; and $0.8$ were $n_j=745; 159;$ and $100$ respectively. See Appendix A.3 for further discussion.
